Replication on multicore considering hierarchy and characteristics of the network

### 2013-04-29 paper.tex
Restructured Section on how to deduce the tree from the network model and added a couple of new ideas

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Building the tree}

First, we build the broadcast tree. We kick-start our protocol
connecting every process with all other processes to get a
fully-meshed network of channels.

@@

% Our implementation
We now build an atomic broadcast for multicore machines. Our
implementation is based on the Barrelfish UMP interconnect driver. It
provides reliable channels with in-order delivery. Communication
starts at the root node, which acts as a
sequentializer. See~\ref{algo:ab} for details.

\newcommand{\textc}[1]{{\color{gray} {\footnotesize #1}}}
\begin{algorithm}[H]
\SetCommentSty{textc}
\SetKwInOut{Assumptions}{assumptions}
\Assumptions{Underlying communication channel is reliable and in-order}
\SetKwProg{Fn}{Function}{}{end}%
\SetKwFunction{receive}{on\_receive}%
\SetKwFunction{waitchild}{wait\_for\_children}%
\SetKwFunction{send}{send}%
\SetKwFunction{icsend}{send\_bc\_request}%
\SetKwFunction{icsendack}{send\_bc\_ack}%
\SetKwFunction{handlemessages}{handle\_other\_messages}%
  %
  \KwData{List of processes $p$, broadcast tree as graph $(V, E)$}
  \KwResult{Tree based atomic broadcast using a sequentializer}
  % 
  \BlankLine
  \Fn(\tcp*{Receive a message}){\receive{$client$, $m$}}{
    \For(\tcp*{For all children}){$c \leftarrow \{ c: \exists (self, c) \in E \} $}{
      \icsend{c}
    }
    \waitchild{}\;
    \icsendack{$client$}\;
  }
  % 
  \BlankLine
  \Fn(\tcp*{Send a message}){\send{void}}{
    \tcc{Need to wait for acknowledgment before returning to
      caller. Otherwise, sender might see his own request before some
      other request, that the sequentializer decided to handle first}
    \icsend{$V_{root}$}\tcp*{Relay msg (sequentializer)}
    \While{no answer received}{
      \handlemessages{}\tcp*{Otherwise, deadlocks}
    }
  }
  \caption{Atomic broadcast on reliable communication channels}
  \label{algo:ab}
\end{algorithm}

% How to deduct tree from model?
We then use this graph to build a group-communication
tree. Interesting techniques could be: clustering, to find closely
connected subsets, distributed messages across cluster first, and then
within cluster.

Maybe also a minimum spanning-tree (MST) protocol, to find a minimal
graph such that all nodes can be reached. The problem with the MST is
that it does not consider parallelism. While it is true that the
global cost is minimal in such a tree, it does not take into account
that sending in parallel does not increase the time required to reach
every node. %
\stefan{Find an example, the MST could be a path through the graph
  with weights 1, where it might make more sense to use a link with
  cost 2 in the beginning and then send messages on two ``branches''
  of the tree in parallel.}

% Radix degree
If the send cost $t_s$ is smaller than the propagation time $t_p$,
\stefan{round up}%
$d = \frac{t_p}{t_s}$ messages can be send to hide propagation of
messages. Hence, a tree with degree $d$ should be ideal to minimize
latency, but only for an acknowledged protocol.

% Tree -> scheduling
Assuming the tree, there is a \emph{scheduling problems} for the order in
which to send messages to children. The intuition is to send on
long-delay links first to hide the more expensive latencies, as this
link is dominating total send cost.

>>>>

We build our atomic broadcast based on an overlay network connecting
all nodes of a given multicore machine.

As a first step, we evaluate tree based overlay networks. Other
potentially interesting topologies are rings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tree}

We now look at tree based atomic broadcast implementations.

The first challenge is to find efficient spanning trees for arbitrary
machines. These machines are represented by a model, which we
described earlier.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Divide and conquer}

We divide the space into clusters. In every round, we further split up
all cluster into smaller clusters and connect them with the minimal
link.

We find this link by searching for links that have start and
end node in different regions of the most recent split. Out of these
links, we pick the cheapest ones. We ignore links whose start or end
node have a too high degree in the spanning tree already.

Assuming that the choice of node IDs represents the topology to some
extend, splitting them up into distinct clusters avoids contention, as
two separate sub-trees are processed in different areas of the
interconnect network.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Clustering}

Another idea is to employ clustering. One example is to base this on
NUMA domains such that every NUMA domain corresponds to one cluster.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Minimum-Spanning Tree}

We can apply a minimum spanning tree algorithm to the machine model
graph to find a number of links such that these links connect all
nodes at a minimum cost.

We have implemented such an algorithm and found out, that it is not a
good solution in the general case, since it does not consider
parallelism. Naturally, we would like to parallelize send operations
to different nodes, which allows them to send messages in parallel.

\stefan{We have an example for this, where the MST is essentially a
  path through the graph}

An important variable in constructing such a tree is the degree of
nodes. If the send cost $t_s$ is smaller than the propagation time
$t_p$,  $d = \lceil \frac{t_p}{t_s} \rceil$ messages can be send to
hide propagation of messages. Hence, a tree with degree $d$ should be
ideal to minimize latency, but only for an acknowledged protocol.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scheduling}

% Tree -> scheduling
Assuming the tree (or some other kind of overlay network), there is a
\emph{scheduling problems} for the order in which to send messages to
children. The intuition is to send on long-delay links first to hide
the more expensive latencies, as this link is dominating total send
cost.

\stefan{We have experiments showing this}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Example tree}

% Our implementation
Assuming the tree, we now build an atomic broadcast for multicore
machines. Our implementation is based on the Barrelfish UMP
interconnect driver. It provides reliable channels with in-order
delivery. Communication starts at the root node, which acts as a
sequentializer. See~\ref{algo:ab} for details.

\newcommand{\textc}[1]{{\color{gray} {\footnotesize #1}}}
\begin{algorithm}[H]
\SetCommentSty{textc}
\SetKwInOut{Assumptions}{assumptions}
\Assumptions{Underlying communication channel is reliable and in-order}
\SetKwProg{Fn}{Function}{}{end}%
\SetKwFunction{receive}{on\_receive}%
\SetKwFunction{waitchild}{wait\_for\_children}%
\SetKwFunction{send}{send}%
\SetKwFunction{icsend}{send\_bc\_request}%
\SetKwFunction{icsendack}{send\_bc\_ack}%
\SetKwFunction{handlemessages}{handle\_other\_messages}%
  %
  \KwData{List of processes $p$, broadcast tree as graph $(V, E)$}
  \KwResult{Tree based atomic broadcast using a sequentializer}
  % 
  \BlankLine
  \Fn(\tcp*{Receive a message}){\receive{$client$, $m$}}{
    \For(\tcp*{For all children}){$c \leftarrow \{ c: \exists (self, c) \in E \} $}{
      \icsend{c}
    }
    \waitchild{}\;
    \icsendack{$client$}\;
  }
  % 
  \BlankLine
  \Fn(\tcp*{Send a message}){\send{void}}{
    \tcc{Need to wait for acknowledgment before returning to
      caller. Otherwise, sender might see his own request before some
      other request, that the sequentializer decided to handle first}
    \icsend{$V_{root}$}\tcp*{Relay msg (sequentializer)}
    \While{no answer received}{
      \handlemessages{}\tcp*{Otherwise, deadlocks}
    }
  }
  \caption{Atomic broadcast on reliable communication channels}
  \label{algo:ab}

\end{algorithm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Barrelfish implementation}

When implementing the protocol in Barrelfish, we found several
practical problems, which we will discuss briefly in this section.

We kick-start our protocol connecting every process with all other
processes to get a fully-meshed network of channels.

@@
